---
title: "SILSO Daily Sunspot Numbers → MySQL (solar schema)"
author: "David Jackson"
format:
  html:
    toc: true
    code-fold: false
  pdf: default
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This notebook:

1. Downloads the official **SILSO Daily Total Sunspot Number (V2.0)**
2. Parses the fixed-width file
3. Creates a MySQL table in the `solar` schema (if not exists)
4. Upserts the data keyed on `date`
5. Verifies the latest rows

---

# 1. Load Required Libraries

```{r}
suppressPackageStartupMessages({
  library(DBI)
  library(RMariaDB)
  library(readr)
  library(dplyr)
  library(lubridate)
})
```

# 2. Configuration

```{r}
mysql_host <- "localhost"
mysql_user <- "root"
mysql_pass <- "dJj135790"
mysql_port <- 3306
mysql_db   <- "silso"

table_name <- "sunspot_daily_silso"

silso_url <- "https://www.sidc.be/silso/DATA/SN_d_tot_V2.0.txt"
```

## 3. Download and Parse SILSO Daily File

```{r}
read_silso_daily_txt <- function(url) {

  tf <- tempfile(fileext = ".txt")
  download.file(url, tf, mode = "wb", quiet = TRUE)

  fwf <- fwf_positions(
    start = c(1, 6, 9, 12, 22, 26, 33, 37),
    end   = c(4, 7,10, 19, 24, 30, 35, 37),
    col_names = c("year","month","day","dec_date","sn","sd","nobs","prov_flag")
  )

  read_fwf(
    tf,
    col_positions = fwf,
    col_types = cols(
      year = col_integer(),
      month = col_integer(),
      day = col_integer(),
      dec_date = col_double(),
      sn = col_double(),
      sd = col_double(),
      nobs = col_integer(),
      prov_flag = col_character()
    ),
    trim_ws = TRUE,
    progress = FALSE
  ) %>%
    mutate(
      date = as.Date(sprintf("%04d-%02d-%02d", year, month, day)),
      sn = na_if(sn, -1),
      sd = na_if(sd, -1),
      prov_flag = if_else(is.na(prov_flag), "", prov_flag),
      is_provisional = if_else(prov_flag == "*", 1L, 0L)
    ) %>%
    select(
      date, year, month, day,
      dec_date, sn, sd, nobs,
      prov_flag, is_provisional
    ) %>%
    arrange(date)
}

silso_df <- read_silso_daily_txt(silso_url)

head(silso_df)
```

## 4. Connect to MySQL

```{r}
con <- dbConnect(
  RMariaDB::MariaDB(),
  host = mysql_host,
  user = mysql_user,
  password = mysql_pass,
  port = mysql_port,
  dbname = mysql_db
)

con
```

## 5. Create Target Table (If Needed)

```{r}
create_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date` DATE NOT NULL,
   `year` SMALLINT NOT NULL,
   `month` TINYINT NOT NULL,
   `day` TINYINT NOT NULL,
   `dec_date` DOUBLE NULL,
   `sn` DOUBLE NULL,
   `sd` DOUBLE NULL,
   `nobs` INT NULL,
   `prov_flag` CHAR(1) NOT NULL DEFAULT '',
   `is_provisional` TINYINT(1) NOT NULL DEFAULT 0,
   `ingested_at` TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP,
   `source` VARCHAR(64) NOT NULL DEFAULT 'WDC-SILSO SN_d_tot_V2.0',
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, table_name)

dbExecute(con, create_sql)
```

## 6. Upsert Data

```{r}
stage_tbl <- paste0(table_name, "_stage")

dbWriteTable(con, stage_tbl, silso_df, overwrite = TRUE)

upsert_sql <- sprintf("
INSERT INTO `%s`.`%s`
  (`date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`)
SELECT
  `date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`
FROM `%s`
ON DUPLICATE KEY UPDATE
  `year` = VALUES(`year`),
  `month` = VALUES(`month`),
  `day` = VALUES(`day`),
  `dec_date` = VALUES(`dec_date`),
  `sn` = VALUES(`sn`),
  `sd` = VALUES(`sd`),
  `nobs` = VALUES(`nobs`),
  `prov_flag` = VALUES(`prov_flag`),
  `is_provisional` = VALUES(`is_provisional`),
  `ingested_at` = CURRENT_TIMESTAMP;",
mysql_db, table_name, stage_tbl)

dbExecute(con, upsert_sql)

dbExecute(con, sprintf("DROP TABLE `%s`;", stage_tbl))
```

## 7. Verify Latest Rows

```{r}
qry <- sprintf("
SELECT date, sn, sd, nobs, is_provisional
FROM `%s`.`%s`
ORDER BY date DESC
LIMIT 10;",
mysql_db, table_name)

dbGetQuery(con, qry)
```