---
title: "SILSO Daily Sunspot Numbers → MySQL (solar schema)"
author: "David Jackson"
format:
  html:
    toc: true
    code-fold: false
  pdf: default
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This notebook:

1. Downloads the official **SILSO Daily Total Sunspot Number (V2.0)**
2. Parses the fixed-width file
3. Creates MySQL tables in the `silso` schema (if not exists):
   - `daily` — full historical record
   - `hemisphere` — full hemispheric record
   - `daily_rolling_100yr` — most recent 100 years to the day (rebuilt on every run)
4. Upserts the full tables; replaces the rolling 100-year table
5. Verifies the latest rows

---

# 1. Load Required Libraries

```{r,message=FALSE,echo=FALSE}
suppressPackageStartupMessages({
  library(DBI)
  library(RMariaDB)
  library(readr)
  library(tidyverse)
  library(lubridate)
  library(zoo)
  library(pracma)
})
```

# 2. Configuration

```{r}
rm(list=ls())
mysql_host <- "localhost"
mysql_user <- "root"
mysql_pass <- "dJj135790"
mysql_port <- 3333
mysql_db   <- "silso"

## Connect to MySQL
con <- dbConnect(
  RMariaDB::MariaDB(),
  host = mysql_host,
  user = mysql_user,
  password = mysql_pass,
  port = mysql_port,
  dbname = mysql_db
)

table_daily      <- "daily"
table_hemp       <- "hemisphere"
table_rolling    <- "daily_rolling_100yr"   # <-- new rolling window table

silso_daily <- "https://www.sidc.be/silso/DATA/SN_d_tot_V2.0.csv"
silso_hemp  <- "https://www.sidc.be/silso/DATA/SN_d_hem_V2.0.csv"
```

## 3. Download and Parse SILSO Files

```{r}
# --- SN_d_tot_V2.0.csv: daily total sunspot numbers ---
read_silso_daily <- function(url) {
  col_names <- c("year", "month", "day", "dec_date", "sn", "sd", "nobs", "prov_flag")

  read_delim(
    url,
    col_names = col_names,
    col_types = cols(
      year      = col_integer(),
      month     = col_integer(),
      day       = col_integer(),
      dec_date  = col_double(),
      sn        = col_double(),
      sd        = col_double(),
      nobs      = col_integer(),
      prov_flag = col_integer()  # 1 = definitive, 0 = provisional
    ),
    delim    = ";",
    trim_ws  = TRUE,
    progress = FALSE
  ) %>%
    mutate(
      date           = as.Date(sprintf("%04d-%02d-%02d", year, month, day)),
      sn             = na_if(sn, -1),
      sd             = na_if(sd, -1),
      is_provisional = if_else(is.na(prov_flag) | prov_flag == 0L, 1L, 0L)
    ) %>%
    select(date, year, month, day, dec_date, sn, sd, nobs, prov_flag, is_provisional) %>%
    arrange(date)
}

# --- SN_d_hem_V2.0.csv: daily hemispheric sunspot numbers ---
read_silso_hemp <- function(url) {
  col_names <- c("year", "month", "day", "dec_date",
                 "sn_tot", "sn_n", "sn_s",
                 "sd_tot", "sd_n", "sd_s",
                 "nobs_tot", "nobs_n", "nobs_s",
                 "prov_flag")

  read_delim(
    url,
    col_names = col_names,
    col_types = cols(
      year      = col_integer(),
      month     = col_integer(),
      day       = col_integer(),
      dec_date  = col_double(),
      sn_tot    = col_double(),
      sn_n      = col_double(),
      sn_s      = col_double(),
      sd_tot    = col_double(),
      sd_n      = col_double(),
      sd_s      = col_double(),
      nobs_tot  = col_integer(),
      nobs_n    = col_integer(),
      nobs_s    = col_integer(),
      prov_flag = col_integer()  # 1 = definitive, 0 = provisional
    ),
    delim    = ";",
    trim_ws  = TRUE,
    progress = FALSE
  ) %>%
    mutate(
      date           = as.Date(sprintf("%04d-%02d-%02d", year, month, day)),
      sn_tot         = na_if(sn_tot, -1),
      sn_n           = na_if(sn_n,   -1),
      sn_s           = na_if(sn_s,   -1),
      sd_tot         = na_if(sd_tot, -1),
      sd_n           = na_if(sd_n,   -1),
      sd_s           = na_if(sd_s,   -1),
      nobs_n         = na_if(nobs_n, -1L),  # always -1 per SILSO docs
      nobs_s         = na_if(nobs_s, -1L),  # always -1 per SILSO docs
      is_provisional = if_else(is.na(prov_flag) | prov_flag == 0L, 1L, 0L)
    ) %>%
    select(date, year, month, day, dec_date,
           sn_tot, sn_n, sn_s,
           sd_tot, sd_n, sd_s,
           nobs_tot, nobs_n, nobs_s,
           prov_flag, is_provisional) %>%
    arrange(date)
}

silso_df      <- read_silso_daily(silso_daily)
silso_hemp_df <- read_silso_hemp(silso_hemp)

head(silso_df)
head(silso_hemp_df)
```

## 3a. Build Rolling 100-Year Slice

The cutoff is computed as today's date minus exactly 100 years (to the day), so
the window automatically advances each time the notebook runs.

```{r}
# Exact 100-year window: same month/day, 100 years back
cutoff_date     <- today() - years(100)
rolling_100_df  <- silso_df %>% filter(date >= cutoff_date)

cat(sprintf(
  "Rolling window: %s → %s  (%d rows)\n",
  min(rolling_100_df$date),
  max(rolling_100_df$date),
  nrow(rolling_100_df)
))
```

## Monthly EDA

```{r}
silso_df$month <- as.factor(silso_df$month)
ggplot(silso_df) +
  geom_boxplot(aes(x = sn, group = month, y = month)) +
  labs(title = "Daily Sunspot Number by Month (full record)",
       x = "Sunspot Number", y = "Month")
```

```{r}
silso_hemp_df$month <- as.factor(silso_hemp_df$month)
ggplot(silso_hemp_df) +
  geom_boxplot(aes(x = sn_n, group = month, y = month)) +
  labs(title = "North Sunspot Number by Month (full record)",
       x = "Sunspot Number", y = "Month")
```
```{r}
silso_hemp_df$month <- as.factor(silso_hemp_df$month)
ggplot(silso_hemp_df) +
  geom_boxplot(aes(x = sn_s, group = month, y = month)) +
  labs(title = "South Sunspot Number by Month (full record)",
       x = "Sunspot Number", y = "Month")
```

## 5. Create Target Tables (If Needed)

```{r}
# --- daily (SN_d_tot_V2.0) ---
create_daily_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date`           DATE NOT NULL,
   `year`           SMALLINT NOT NULL,
   `month`          TINYINT NOT NULL,
   `day`            TINYINT NOT NULL,
   `dec_date`       DOUBLE NULL,
   `sn`             DOUBLE NULL,
   `sd`             DOUBLE NULL,
   `nobs`           INT NULL,
   `prov_flag`      TINYINT(1) NOT NULL DEFAULT 1,
   `is_provisional` TINYINT(1) NOT NULL DEFAULT 0,
   `ingested_at`    TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP,
   `source`         VARCHAR(64) NOT NULL DEFAULT 'WDC-SILSO SN_d_tot_V2.0',
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, table_daily)

# --- hemisphere (SN_d_hem_V2.0) ---
create_hemp_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date`           DATE NOT NULL,
   `year`           SMALLINT NOT NULL,
   `month`          TINYINT NOT NULL,
   `day`            TINYINT NOT NULL,
   `dec_date`       DOUBLE NULL,
   `sn_tot`         DOUBLE NULL,
   `sn_n`           DOUBLE NULL,
   `sn_s`           DOUBLE NULL,
   `sd_tot`         DOUBLE NULL,
   `sd_n`           DOUBLE NULL,
   `sd_s`           DOUBLE NULL,
   `nobs_tot`       INT NULL,
   `nobs_n`         INT NULL,
   `nobs_s`         INT NULL,
   `prov_flag`      TINYINT(1) NOT NULL DEFAULT 1,
   `is_provisional` TINYINT(1) NOT NULL DEFAULT 0,
   `ingested_at`    TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP,
   `source`         VARCHAR(64) NOT NULL DEFAULT 'WDC-SILSO SN_d_hem_V2.0',
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, table_hemp)

# --- daily_rolling_100yr ---
# Same structure as `daily`; rebuilt completely on every run so no upsert needed.
# window_start / window_end columns record what range was loaded.
create_rolling_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date`           DATE NOT NULL,
   `year`           SMALLINT NOT NULL,
   `month`          TINYINT NOT NULL,
   `day`            TINYINT NOT NULL,
   `dec_date`       DOUBLE NULL,
   `sn`             DOUBLE NULL,
   `sd`             DOUBLE NULL,
   `nobs`           INT NULL,
   `prov_flag`      TINYINT(1) NOT NULL DEFAULT 1,
   `is_provisional` TINYINT(1) NOT NULL DEFAULT 0,
   `window_start`   DATE NOT NULL COMMENT 'Earliest date in this 100-yr load',
   `window_end`     DATE NOT NULL COMMENT 'Latest date in this 100-yr load',
   `ingested_at`    TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP,
   `source`         VARCHAR(64) NOT NULL DEFAULT 'WDC-SILSO SN_d_tot_V2.0',
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, table_rolling)

dbExecute(con, create_daily_sql)
dbExecute(con, create_hemp_sql)
dbExecute(con, create_rolling_sql)
```

## 6. Insert Data

```{r}
# --- Upsert: daily ---
stage_daily <- paste0(table_daily, "_stage")
dbWriteTable(con, stage_daily, silso_df, overwrite = TRUE)

upsert_daily_sql <- sprintf("
INSERT INTO `%s`.`%s`
  (`date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`)
SELECT
  `date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`
FROM `%s`
ON DUPLICATE KEY UPDATE
  `dec_date`       = VALUES(`dec_date`),
  `sn`             = VALUES(`sn`),
  `sd`             = VALUES(`sd`),
  `nobs`           = VALUES(`nobs`),
  `prov_flag`      = VALUES(`prov_flag`),
  `is_provisional` = VALUES(`is_provisional`),
  `ingested_at`    = CURRENT_TIMESTAMP;",
mysql_db, table_daily, stage_daily)

dbExecute(con, upsert_daily_sql)
dbExecute(con, sprintf("DROP TABLE `%s`;", stage_daily))

# --- Upsert: hemisphere ---
stage_hemp <- paste0(table_hemp, "_stage")
dbWriteTable(con, stage_hemp, silso_hemp_df, overwrite = TRUE)

upsert_hemp_sql <- sprintf("
INSERT INTO `%s`.`%s`
  (`date`,`year`,`month`,`day`,`dec_date`,
   `sn_tot`,`sn_n`,`sn_s`,
   `sd_tot`,`sd_n`,`sd_s`,
   `nobs_tot`,`nobs_n`,`nobs_s`,
   `prov_flag`,`is_provisional`)
SELECT
  `date`,`year`,`month`,`day`,`dec_date`,
  `sn_tot`,`sn_n`,`sn_s`,
  `sd_tot`,`sd_n`,`sd_s`,
  `nobs_tot`,`nobs_n`,`nobs_s`,
  `prov_flag`,`is_provisional`
FROM `%s`
ON DUPLICATE KEY UPDATE
  `dec_date`       = VALUES(`dec_date`),
  `sn_tot`         = VALUES(`sn_tot`),
  `sn_n`           = VALUES(`sn_n`),
  `sn_s`           = VALUES(`sn_s`),
  `sd_tot`         = VALUES(`sd_tot`),
  `sd_n`           = VALUES(`sd_n`),
  `sd_s`           = VALUES(`sd_s`),
  `nobs_tot`       = VALUES(`nobs_tot`),
  `nobs_n`         = VALUES(`nobs_n`),
  `nobs_s`         = VALUES(`nobs_s`),
  `prov_flag`      = VALUES(`prov_flag`),
  `is_provisional` = VALUES(`is_provisional`),
  `ingested_at`    = CURRENT_TIMESTAMP;",
mysql_db, table_hemp, stage_hemp)

dbExecute(con, upsert_hemp_sql)
dbExecute(con, sprintf("DROP TABLE `%s`;", stage_hemp))
```

## 6a. Load Rolling 100-Year Table

The rolling table is **fully replaced** on every run — no upsert needed — so the
window boundaries are always current.

```{r}
# Annotate with window metadata before writing
window_start <- min(rolling_100_df$date)
window_end   <- max(rolling_100_df$date)

rolling_load_df <- rolling_100_df %>%
  mutate(
    window_start = window_start,
    window_end   = window_end
  )

# Truncate then reload: ensures stale rows at the old leading edge are purged
dbExecute(con, sprintf("TRUNCATE TABLE `%s`.`%s`;", mysql_db, table_rolling))

dbWriteTable(
  con,
  name      = table_rolling,
  value     = rolling_load_df,
  overwrite = FALSE,   # table already exists; just append into the empty table
  append    = TRUE
)

cat(sprintf(
  "Loaded %d rows into %s.%s\n  window_start : %s\n  window_end   : %s\n",
  nrow(rolling_load_df), mysql_db, table_rolling,
  window_start, window_end
))
```

## 7. Verify Latest Rows

```{r}
# daily
dbGetQuery(con, sprintf("
SELECT date, sn, sd, nobs, is_provisional
FROM `%s`.`%s`
ORDER BY date DESC LIMIT 10;",
mysql_db, table_daily))

# hemisphere
dbGetQuery(con, sprintf("
SELECT date, sn_tot, sn_n, sn_s, is_provisional
FROM `%s`.`%s`
ORDER BY date DESC LIMIT 10;",
mysql_db, table_hemp))

# rolling 100-year — newest 10 and oldest 10 to confirm window edges
cat("--- Rolling 100yr: newest rows ---\n")
dbGetQuery(con, sprintf("
SELECT date, sn, sd, nobs, window_start, window_end, is_provisional
FROM `%s`.`%s`
ORDER BY date DESC LIMIT 10;",
mysql_db, table_rolling))

cat("--- Rolling 100yr: oldest rows ---\n")
dbGetQuery(con, sprintf("
SELECT date, sn, sd, nobs, window_start, window_end, is_provisional
FROM `%s`.`%s`
ORDER BY date ASC LIMIT 10;",
mysql_db, table_rolling))

# Row count sanity check
dbGetQuery(con, sprintf("
SELECT
  COUNT(*)           AS total_rows,
  MIN(date)          AS earliest,
  MAX(date)          AS latest,
  DATEDIFF(MAX(date), MIN(date)) AS span_days
FROM `%s`.`%s`;",
mysql_db, table_rolling))
```
