---
title: "SILSO Daily Sunspot Numbers → MySQL (solar schema)"
author: "David Jackson"
format:
  html:
    toc: true
    code-fold: false
  pdf: default
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This notebook:

1. Downloads the official **SILSO Daily Total Sunspot Number (V2.0)**
2. Parses the fixed-width file
3. Creates a MySQL table in the `solar` schema (if not exists)
4. Upserts the data keyed on `date`
5. Verifies the latest rows

---

# 1. Load Required Libraries

```{r}
suppressPackageStartupMessages({
  library(DBI)
  library(RMariaDB)
  library(readr)
  library(tidyverse)
  library(lubridate)
})
```

# 2. Configuration

```{r}
mysql_host <- "localhost"
mysql_user <- "root"
mysql_pass <- "dJj135790"
mysql_port <- 3333
mysql_db   <- "silso"


## 4. Connect to MySQL

con <- dbConnect(
  RMariaDB::MariaDB(),
  host = mysql_host,
  user = mysql_user,
  password = mysql_pass,
  port = mysql_port,
  dbname = mysql_db
)



table_name <- "daily"

silso_daily <-"https://www.sidc.be/silso/DATA/SN_d_tot_V2.0.csv"
silso_hemp <-"https://www.sidc.be/silso/DATA/SN_d_hem_V2.0.csv"
```

## 3. Download and Parse SILSO Daily File

```{r}
read_silso_csv <- function(url, col_names) {
  read_delim(
    url,
    col_names = col_names,
    col_types = cols(
      year       = col_integer(),
      month      = col_integer(),
      day        = col_integer(),
      dec_date   = col_double(),
      sn         = col_double(),
      sd         = col_double(),
      nobs       = col_integer(),
      prov_flag  = col_integer()   # CSV encodes: 1 = definitive, 0 = provisional
    ),
    delim     = ";",
    trim_ws   = TRUE,
    progress  = FALSE
  ) %>%
    mutate(
      date = as.Date(sprintf("%04d-%02d-%02d", year, month, day)),
      sn = na_if(sn, -1),
      sd = na_if(sd, -1),
      # CSV: 1 = definitive, 0 = provisional  →  invert to is_provisional flag
      is_provisional = if_else(is.na(prov_flag) | prov_flag == 0L, 1L, 0L)
    ) %>%
    arrange(date)
}

# Column names for the total sunspot number file (SN_d_tot_V2.0.csv)
daily_cols <- c("year", "month", "day", "dec_date", "sn", "sd", "nobs", "prov_flag")

# Column names for the hemispheric file (SN_d_hem_V2.0.csv)
# 14 columns per SILSO documentation:
#   date (3) | dec_date | sn_tot, sn_n, sn_s | sd_tot, sd_n, sd_s |
#   nobs_tot, nobs_n, nobs_s | prov_flag
hemp_cols  <- c("year", "month", "day", "dec_date",
                "sn_tot", "sn_n", "sn_s",
                "sd_tot", "sd_n", "sd_s",
                "nobs_tot", "nobs_n", "nobs_s",
                "prov_flag")

silso_df      <- read_silso_csv(silso_daily, daily_cols) %>%
  select(date, year, month, day, dec_date, sn, sd, nobs, prov_flag, is_provisional)

silso_hemp_df <- read_silso_csv(silso_hemp, hemp_cols) %>%
  select(date, year, month, day, dec_date,
         sn_tot, sn_n, sn_s,
         sd_tot, sd_n, sd_s,
         nobs_tot, nobs_n, nobs_s,
         prov_flag, is_provisional)

head(silso_df)
head(silso_hemp_df)
```
## Monthly EDA

```{r}
silso_df$month <- as.factor(silso_df$month)
ggplot(silso_df) + geom_boxplot(aes(x=sn,group=month,y=month))
```





## 5. Create Target Table (If Needed)

```{r}
create_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date` DATE NOT NULL,
   `year` SMALLINT NOT NULL,
   `month` TINYINT NOT NULL,
   `day` TINYINT NOT NULL,
   `dec_date` DOUBLE NULL,
   `sn` DOUBLE NULL,
   `sd` DOUBLE NULL,
   `nobs` INT NULL,
   `prov_flag` CHAR(1) NOT NULL DEFAULT '',
   `is_provisional` TINYINT(1) NOT NULL DEFAULT 0,
   `ingested_at` TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP,
   `source` VARCHAR(64) NOT NULL DEFAULT 'WDC-SILSO SN_d_tot_V2.0',
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, table_name)

dbExecute(con, create_sql)
```

## 6. Upsert Data

```{r}
stage_tbl <- paste0(table_name, "_stage")

dbWriteTable(con, stage_tbl, silso_df, overwrite = TRUE)

upsert_sql <- sprintf("
INSERT INTO `%s`.`%s`
  (`date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`)
SELECT
  `date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`
FROM `%s`
ON DUPLICATE KEY UPDATE
  `year` = VALUES(`year`),
  `month` = VALUES(`month`),
  `day` = VALUES(`day`),
  `dec_date` = VALUES(`dec_date`),
  `sn` = VALUES(`sn`),
  `sd` = VALUES(`sd`),
  `nobs` = VALUES(`nobs`),
  `prov_flag` = VALUES(`prov_flag`),
  `is_provisional` = VALUES(`is_provisional`),
  `ingested_at` = CURRENT_TIMESTAMP;",
mysql_db, table_name, stage_tbl)

dbExecute(con, upsert_sql)

dbExecute(con, sprintf("DROP TABLE `%s`;", stage_tbl))
```

## 7. Verify Latest Rows

```{r}
qry <- sprintf("
SELECT date, sn, sd, nobs, is_provisional
FROM `%s`.`%s`
ORDER BY date DESC
LIMIT 10;",
mysql_db, table_name)

dbGetQuery(con, qry)
```

## Begin 100 Year Rolling Period


```{r}
rolling_table_name <- "sunspot_daily_silso_100y"

create_rolling_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date` DATE NOT NULL,
   `year` SMALLINT NOT NULL,
   `month` TINYINT NOT NULL,
   `day` TINYINT NOT NULL,
   `dec_date` DOUBLE NULL,
   `sn` DOUBLE NULL,
   `sd` DOUBLE NULL,
   `nobs` INT NULL,
   `prov_flag` CHAR(1) NOT NULL DEFAULT '',
   `is_provisional` TINYINT(1) NOT NULL DEFAULT 0,
   `refreshed_at` TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP
     ON UPDATE CURRENT_TIMESTAMP,
   `source_table` VARCHAR(64) NOT NULL DEFAULT 'sunspot_daily_silso',
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, rolling_table_name)

dbExecute(con, create_rolling_sql)
```

## 3) Refresh it each run (simple + safe)

```{r}
# Step 3 (fixed): refresh the rolling 100-year table

# 1) TRUNCATE as its own statement
truncate_sql <- sprintf(
  "TRUNCATE TABLE `%s`.`%s`",
  mysql_db, rolling_table_name
)
dbExecute(con, truncate_sql)

# 2) INSERT as its own statement
insert_sql <- sprintf("
  INSERT INTO `%s`.`%s`
    (`date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`)
  SELECT
    `date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`
  FROM `%s`.`%s`
  WHERE `date` >= (CURDATE() - INTERVAL 100 YEAR)
  ORDER BY `date`
",
  mysql_db, rolling_table_name,
  mysql_db, table_name
)

dbExecute(con, insert_sql)
```
