---
title: "SILSO Daily Sunspot Numbers → MySQL (solar schema)"
author: "David Jackson"
format:
  html:
    toc: true
    code-fold: false
  pdf: default
execute:
  echo: true
  warning: false
  message: false
---

# Overview

This notebook:

1. Downloads the official **SILSO Daily Total Sunspot Number (V2.0)**
2. Parses the fixed-width file
3. Creates a MySQL table in the `solar` schema (if not exists)
4. Upserts the data keyed on `date`
5. Verifies the latest rows

---

# 1. Load Required Libraries

```{r}
suppressPackageStartupMessages({
  library(DBI)
  library(RMariaDB)
  library(readr)
  library(tidyverse)
  library(lubridate)
  library(zoo)
  library(pracma)
})
```

# 2. Configuration

```{r}
mysql_host <- "localhost"
mysql_user <- "root"
mysql_pass <- "dJj135790"
mysql_port <- 3333
mysql_db   <- "silso"


## 4. Connect to MySQL


con <- dbConnect(
  RMariaDB::MariaDB(),
  host = mysql_host,
  user = mysql_user,
  password = mysql_pass,
  port = mysql_port,
  dbname = mysql_db
)

con


table_daily <- "daily"
table_hemp  <- "hemisphere"

silso_daily <-"https://www.sidc.be/silso/DATA/SN_d_tot_V2.0.csv"
silso_hemp <-"https://www.sidc.be/silso/DATA/SN_d_hem_V2.0.csv"
```

## 3. Download and Parse SILSO Files

```{r}
# --- SN_d_tot_V2.0.csv: daily total sunspot numbers ---
read_silso_daily <- function(url) {
  col_names <- c("year", "month", "day", "dec_date", "sn", "sd", "nobs", "prov_flag")

  read_delim(
    url,
    col_names = col_names,
    col_types = cols(
      year      = col_integer(),
      month     = col_integer(),
      day       = col_integer(),
      dec_date  = col_double(),
      sn        = col_double(),
      sd        = col_double(),
      nobs      = col_integer(),
      prov_flag = col_integer()  # 1 = definitive, 0 = provisional
    ),
    delim    = ";",
    trim_ws  = TRUE,
    progress = FALSE
  ) %>%
    mutate(
      date           = as.Date(sprintf("%04d-%02d-%02d", year, month, day)),
      sn             = na_if(sn, -1),
      sd             = na_if(sd, -1),
      is_provisional = if_else(is.na(prov_flag) | prov_flag == 0L, 1L, 0L)
    ) %>%
    select(date, year, month, day, dec_date, sn, sd, nobs, prov_flag, is_provisional) %>%
    arrange(date)
}

# --- SN_d_hem_V2.0.csv: daily hemispheric sunspot numbers ---
read_silso_hemp <- function(url) {
  col_names <- c("year", "month", "day", "dec_date",
                 "sn_tot", "sn_n", "sn_s",
                 "sd_tot", "sd_n", "sd_s",
                 "nobs_tot", "nobs_n", "nobs_s",
                 "prov_flag")

  read_delim(
    url,
    col_names = col_names,
    col_types = cols(
      year     = col_integer(),
      month    = col_integer(),
      day      = col_integer(),
      dec_date = col_double(),
      sn_tot   = col_double(),
      sn_n     = col_double(),
      sn_s     = col_double(),
      sd_tot   = col_double(),
      sd_n     = col_double(),
      sd_s     = col_double(),
      nobs_tot = col_integer(),
      nobs_n   = col_integer(),
      nobs_s   = col_integer(),
      prov_flag = col_integer()  # 1 = definitive, 0 = provisional
    ),
    delim    = ";",
    trim_ws  = TRUE,
    progress = FALSE
  ) %>%
    mutate(
      date           = as.Date(sprintf("%04d-%02d-%02d", year, month, day)),
      sn_tot         = na_if(sn_tot, -1),
      sn_n           = na_if(sn_n,   -1),
      sn_s           = na_if(sn_s,   -1),
      sd_tot         = na_if(sd_tot, -1),
      sd_n           = na_if(sd_n,   -1),
      sd_s           = na_if(sd_s,   -1),
      nobs_n         = na_if(nobs_n, -1L),  # always -1 per SILSO docs
      nobs_s         = na_if(nobs_s, -1L),  # always -1 per SILSO docs
      is_provisional = if_else(is.na(prov_flag) | prov_flag == 0L, 1L, 0L)
    ) %>%
    select(date, year, month, day, dec_date,
           sn_tot, sn_n, sn_s,
           sd_tot, sd_n, sd_s,
           nobs_tot, nobs_n, nobs_s,
           prov_flag, is_provisional) %>%
    arrange(date)
}

silso_df      <- read_silso_daily(silso_daily)
silso_hemp_df <- read_silso_hemp(silso_hemp)

head(silso_df)
head(silso_hemp_df)
```
## Monthly EDA

```{r}
silso_df$month <- as.factor(silso_df$month)
ggplot(silso_df) + geom_boxplot(aes(x=sn,group=month,y=month))
```





## 5. Create Target Tables (If Needed)

```{r}
# --- daily (SN_d_tot_V2.0) ---
create_daily_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date`           DATE NOT NULL,
   `year`           SMALLINT NOT NULL,
   `month`          TINYINT NOT NULL,
   `day`            TINYINT NOT NULL,
   `dec_date`       DOUBLE NULL,
   `sn`             DOUBLE NULL,
   `sd`             DOUBLE NULL,
   `nobs`           INT NULL,
   `prov_flag`      TINYINT(1) NOT NULL DEFAULT 1,
   `is_provisional` TINYINT(1) NOT NULL DEFAULT 0,
   `ingested_at`    TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP,
   `source`         VARCHAR(64) NOT NULL DEFAULT 'WDC-SILSO SN_d_tot_V2.0',
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, table_daily)

# --- hemisphere (SN_d_hem_V2.0) ---
create_hemp_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date`           DATE NOT NULL,
   `year`           SMALLINT NOT NULL,
   `month`          TINYINT NOT NULL,
   `day`            TINYINT NOT NULL,
   `dec_date`       DOUBLE NULL,
   `sn_tot`         DOUBLE NULL,
   `sn_n`           DOUBLE NULL,
   `sn_s`           DOUBLE NULL,
   `sd_tot`         DOUBLE NULL,
   `sd_n`           DOUBLE NULL,
   `sd_s`           DOUBLE NULL,
   `nobs_tot`       INT NULL,
   `nobs_n`         INT NULL,
   `nobs_s`         INT NULL,
   `prov_flag`      TINYINT(1) NOT NULL DEFAULT 1,
   `is_provisional` TINYINT(1) NOT NULL DEFAULT 0,
   `ingested_at`    TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP,
   `source`         VARCHAR(64) NOT NULL DEFAULT 'WDC-SILSO SN_d_hem_V2.0',
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, table_hemp)

dbExecute(con, create_daily_sql)
dbExecute(con, create_hemp_sql)
```

## 6. Upsert Data

```{r}
# --- Upsert: daily ---
stage_daily <- paste0(table_daily, "_stage")
dbWriteTable(con, stage_daily, silso_df, overwrite = TRUE)

upsert_daily_sql <- sprintf("
INSERT INTO `%s`.`%s`
  (`date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`)
SELECT
  `date`,`year`,`month`,`day`,`dec_date`,`sn`,`sd`,`nobs`,`prov_flag`,`is_provisional`
FROM `%s`
ON DUPLICATE KEY UPDATE
  `dec_date`       = VALUES(`dec_date`),
  `sn`             = VALUES(`sn`),
  `sd`             = VALUES(`sd`),
  `nobs`           = VALUES(`nobs`),
  `prov_flag`      = VALUES(`prov_flag`),
  `is_provisional` = VALUES(`is_provisional`),
  `ingested_at`    = CURRENT_TIMESTAMP;",
mysql_db, table_daily, stage_daily)

dbExecute(con, upsert_daily_sql)
dbExecute(con, sprintf("DROP TABLE `%s`;", stage_daily))

# --- Upsert: hemisphere ---
stage_hemp <- paste0(table_hemp, "_stage")
dbWriteTable(con, stage_hemp, silso_hemp_df, overwrite = TRUE)

upsert_hemp_sql <- sprintf("
INSERT INTO `%s`.`%s`
  (`date`,`year`,`month`,`day`,`dec_date`,
   `sn_tot`,`sn_n`,`sn_s`,
   `sd_tot`,`sd_n`,`sd_s`,
   `nobs_tot`,`nobs_n`,`nobs_s`,
   `prov_flag`,`is_provisional`)
SELECT
  `date`,`year`,`month`,`day`,`dec_date`,
  `sn_tot`,`sn_n`,`sn_s`,
  `sd_tot`,`sd_n`,`sd_s`,
  `nobs_tot`,`nobs_n`,`nobs_s`,
  `prov_flag`,`is_provisional`
FROM `%s`
ON DUPLICATE KEY UPDATE
  `dec_date`       = VALUES(`dec_date`),
  `sn_tot`         = VALUES(`sn_tot`),
  `sn_n`           = VALUES(`sn_n`),
  `sn_s`           = VALUES(`sn_s`),
  `sd_tot`         = VALUES(`sd_tot`),
  `sd_n`           = VALUES(`sd_n`),
  `sd_s`           = VALUES(`sd_s`),
  `nobs_tot`       = VALUES(`nobs_tot`),
  `nobs_n`         = VALUES(`nobs_n`),
  `nobs_s`         = VALUES(`nobs_s`),
  `prov_flag`      = VALUES(`prov_flag`),
  `is_provisional` = VALUES(`is_provisional`),
  `ingested_at`    = CURRENT_TIMESTAMP;",
mysql_db, table_hemp, stage_hemp)

dbExecute(con, upsert_hemp_sql)
dbExecute(con, sprintf("DROP TABLE `%s`;", stage_hemp))
```

## 7. Verify Latest Rows

```{r}
# daily
dbGetQuery(con, sprintf("
SELECT date, sn, sd, nobs, is_provisional
FROM `%s`.`%s`
ORDER BY date DESC LIMIT 10;",
mysql_db, table_daily))

# hemisphere
dbGetQuery(con, sprintf("
SELECT date, sn_tot, sn_n, sn_s, is_provisional
FROM `%s`.`%s`
ORDER BY date DESC LIMIT 10;",
mysql_db, table_hemp))
```
## Begin 100 Year Rolling Period

```{r}
rolling_daily_name <- "daily_100y"
rolling_hemp_name  <- "hemisphere_100y"

# --- daily_100y ---
create_rolling_daily_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date`         DATE NOT NULL,
   `year`         SMALLINT NOT NULL,
   `month`        TINYINT NOT NULL,
   `day`          TINYINT NOT NULL,
   `dec_date`     DOUBLE NULL,
   `sn_ma100`     DOUBLE NULL COMMENT '100-year rolling mean of daily total sunspot number',
   `sd_ma100`     DOUBLE NULL COMMENT '100-year rolling mean of daily std deviation',
   `refreshed_at` TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP
     ON UPDATE CURRENT_TIMESTAMP,
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, rolling_daily_name)

# --- hemisphere_100y ---
create_rolling_hemp_sql <- sprintf(
"CREATE TABLE IF NOT EXISTS `%s`.`%s` (
   `date`           DATE NOT NULL,
   `year`           SMALLINT NOT NULL,
   `month`          TINYINT NOT NULL,
   `day`            TINYINT NOT NULL,
   `dec_date`       DOUBLE NULL,
   `sn_tot_ma100`   DOUBLE NULL COMMENT '100-year rolling mean of total sunspot number',
   `sn_n_ma100`     DOUBLE NULL COMMENT '100-year rolling mean of North hemisphere sunspot number',
   `sn_s_ma100`     DOUBLE NULL COMMENT '100-year rolling mean of South hemisphere sunspot number',
   `refreshed_at`   TIMESTAMP NOT NULL DEFAULT CURRENT_TIMESTAMP
     ON UPDATE CURRENT_TIMESTAMP,
   PRIMARY KEY (`date`),
   KEY `idx_year_month` (`year`,`month`)
) ENGINE=InnoDB DEFAULT CHARSET=utf8mb4;",
mysql_db, rolling_hemp_name)

dbExecute(con, create_rolling_daily_sql)
dbExecute(con, create_rolling_hemp_sql)
```

## Compute and Write 100-Year Rolling Averages

A 100-year rolling average uses a window of 36,524 days (~100 * 365.24).
`pracma::movavg` is used with `type = "s"` (simple moving average), which
returns a full-length vector with the leading values computed on partial
windows as the series builds up.

```{r}
library(pracma)

window_days <- 36524L  # 100 years * 365.24 days

# --- daily_100y: rolling mean of sn and sd ---
daily_100y_df <- silso_df %>%
  arrange(date) %>%
  mutate(
    sn_ma100 = movavg(replace(sn, is.na(sn), 0), n = window_days, type = "s"),
    sd_ma100 = movavg(replace(sd, is.na(sd), 0), n = window_days, type = "s")
  ) %>%
  select(date, year, month, day, dec_date, sn_ma100, sd_ma100)

# --- hemisphere_100y: rolling mean of sn_tot, sn_n, sn_s ---
hemp_100y_df <- silso_hemp_df %>%
  arrange(date) %>%
  mutate(
    sn_tot_ma100 = movavg(replace(sn_tot, is.na(sn_tot), 0), n = window_days, type = "s"),
    sn_n_ma100   = movavg(replace(sn_n,   is.na(sn_n),   0), n = window_days, type = "s"),
    sn_s_ma100   = movavg(replace(sn_s,   is.na(sn_s),   0), n = window_days, type = "s")
  ) %>%
  select(date, year, month, day, dec_date,
         sn_tot_ma100, sn_n_ma100, sn_s_ma100)
```

```{r}
# --- Write daily_100y to MySQL ---
dbExecute(con, sprintf("TRUNCATE TABLE `%s`.`%s`", mysql_db, rolling_daily_name))
dbWriteTable(con, rolling_daily_name, daily_100y_df, overwrite = FALSE, append = TRUE)

# --- Write hemisphere_100y to MySQL ---
dbExecute(con, sprintf("TRUNCATE TABLE `%s`.`%s`", mysql_db, rolling_hemp_name))
dbWriteTable(con, rolling_hemp_name, hemp_100y_df, overwrite = FALSE, append = TRUE)
```

```{r}
# --- Verify ---
dbGetQuery(con, sprintf("
SELECT date, sn_ma100, sd_ma100
FROM `%s`.`%s`
ORDER BY date DESC LIMIT 10;",
mysql_db, rolling_daily_name))

dbGetQuery(con, sprintf("
SELECT date, sn_tot_ma100, sn_n_ma100, sn_s_ma100
FROM `%s`.`%s`
ORDER BY date DESC LIMIT 10;",
mysql_db, rolling_hemp_name))
```
